Utilization of Reductions and Abstraction Elimination in Typed Genetic Programming

### 2014-04-15 tex/gecco14_old/gecco14.tex
cla

In our system we use a simply typed lambda calculus. Since abstraction elimination and reductions do not involve many type related issues it seems to us as a sufficient choice of a type system. Techniques described here would be analogous for a more sophisticated type system and can be directly applied to such systems as the Hindley–Milner type system.

In this paper we present an approach to typed GP over lambda calculus capable of operating with the full array of notions given by a simply typed lambda calculus while being a direct generalization of standard GP~\cite{koza92}. We discuss the advantages of generating lambda terms in their long normal form and present two novel crossover operations for simply typed lambda terms. Compared to the approach to typed GP over lambda calculus  proposed by 
Yu~\cite{yu01}, it uses more general term representation which is not restricting the use of outer local variables in the body of an anonymous function. Compared to the purely combinator approach by Briggs and 
O’Neill~\cite{kes}, it has the advantage of reducing the size of the search space in the term generation phase.

@@

The rest of the paper is organized as follows: The next section briefly discusses related work in the field of typed GP, while section~\ref{preliminaries} introduces necessary notions.
The body of the paper is presented in sections \ref{generating} and \ref{crossover}; the former discusses the term generating 
procedure and its relations
with reductions, the latter describes several approaches to lambda term
crossover, and how they solve the variable problem. Section~\ref{experiments} examines those ideas on the even parity benchmark problem, while the paper is concluded by section~\ref{conclusions}.

@@

Yu presents a GP system utilizing
polymorphic higher-order functions
%\footnote{Higher-order function is a function taking another function as input parameter.} 
and lambda abstractions~\cite{yu01}.
An important point of interest in the work is the use of
\texttt{foldr} function as a tool for \textit{implicit recursion}, i.e., recursion without explicit recursive calls. The terminal set for constructing lambda abstraction subtrees  is limited to use only constants and variables of that particular lambda abstraction, i.e., outer variables are not allowed to be used as terminals in this work. This is a significant difference from our approach since we permit all well-typed normalized \lterms. This difference also causes the need for different crossover operators.

Briggs and O’Neill present a technique 
utilizing typed GP with combinators \cite{kes}.
The difference between the approach presented in their work
and our approach is that they generate terms in a straightforward way directly from the library of combinators, without any use of lambda abstractions. They use the Hindley–Milner type system. They also present an interesting concept of \textit{Generalized genetic operator} based on term generation. The combinator based genetic algorithm has the advantage of avoiding the need of dealing with variables.

Binard and Felty use an even stronger type system, the (\textit{System F}) \cite{binard2008genetic}. But with the increasing power of the type system also comes an increasing difficulty of term generation. For this reason, the evolution in this work takes interesting and nonstandard shape (fitness is associated with \textit{genes} which are evolved together with \textit{species} which together participate in creation of individuals). This differs from our approach which tries to be a straightforward generalization of the standard GP\cite{koza92}.

In contrast to the above mentioned works, our approach uses a very simple type system (the simply typed lambda calculus) and concentrates on process of generation  
able to generate all possible well-typed normalized lambda terms. In order to do so we use technique based on \textit{inhabitation machines} described by Barendregt\cite{barendregt10}.

@@

Each experiment consisted of 50 independent runs 
of GP algorithm. Each run had a limit of 51 generations, and the population size
was 500 individuals.
In the experiments we analyze the fitness of the best individual,
the average size of an individual,
the ability of the system to produce a correct solution, 
and computational cost estimates throughout generations. 
For the last two metrics we use the popular measurement  
methods within the GP field --- the \textit{performance curves}
described in \cite{koza92} --- $P(M,i)$ (cumulative probability of success) 
and $I(M,i,z)$ (the total number of individuals that must be processed 
to yield a correct solution with probability $z =99\%$).
Difference of those two crossover methods is statistically analyzed by the Welch t-test by comparing the fitness values of the best individuals of the run. 
Graphs showing mean values of the fitness of the best individual and
the average size of an individual throughout generations contain error bars representing the standard error of the mean (SEM).

@@

The \textit{unpacking} crossover scored 23/50 (46\%) success rate. 
Minimal $I(M,i,z)$ was in generation 0 with 114,000 individuals to be processed.
The average individual size for generation 50 was 83.1.

The \textit{hybrid} crossover scored 32/50 (64\%) success rate. 
Minimal $I(M,i,z)$ was in generation 0 with 28,000 individuals to be processed.
The average individual size for generation 50 was 74.3.

Comparison with other results taken from literature is summarized in table 2.
Concerning the time complexity, both experiments were run under same conditions on a fairly standard machine with 4 core 3.3 GHz processor and 4 GB RAM. The first experiment took 388 minutes while the second one took 
33 minutes.

@@

We have also analyzed ratio of failed crossovers due to exceeding the maximal permissible depth. Only around 0.5\% of offspring created by the unpacking crossover fail due to exceeding the limit and 0.2\% for the hybrid crossover. Thus we conclude that the case of crossover failure does not have a significant effect on the performance of the operators.

@@

\begin{figure}[!ht]
  \centering
  \caption{Hybrid crossover example.}
  \includegraphics[scale=0.59]{crossover.png} %0.251
\end{figure}

@@

The first one --- the \textit{hybrid} crossover --- transforms all generated individuals into combinator terms at the end of generation phase. The second crossover approach --- the \textit{unpacking} crossover --- uses the lambda term representation during the whole run, except for the temporary phases during the crossover. We have compared their relative performance on even parity problem; the first one is performing the same or better in all observed metrics. Since the use of the \textit{unpacking} crossover brings a significant computational overhead while its performance is the same or worse than the performance of the \textit{hybrid} crossover we conclude that 
the \textit{hybrid} crossover is better suited for the job.
%\red{This was surprising result for us.}

According to comparison of "individuals to be processed" metric our GP approach is performing better than the GP system presented in \cite{kes} and worse than the one presented in \cite{yu01}. By observation of typical solution which is often some modification of \texttt{foldr xor True inputList} one sees that an important task for GP here is to create \texttt{xor} function. This is the advantage for the more specialized term representation used in \cite{yu01}, which restricts the use of outer variables.

@@

Tom\'{a}\v{s} K\v{r}en has been partially supported by the project GA \v{C}R P202/10/1333 and by the SVV project number 260~104. Roman Neruda has been partially supported by The Ministry of Education of the Czech Republic project COST LD 13002.

>>>>

In our system we use simply typed lambda calculus. Since abstraction elimination and reductions do not involve many type related issues it seems to us as a sufficient choice of type system. Techniques described here would be analogous for a more sophisticated type system and can be directly applied to such systems as the Hindley–Milner type system.

In this paper we present an approach to typed GP over lambda calculus capable of operating with the full arsenal given by simply typed lambda calculus while being a direct generalization of standard GP\cite{koza92}. We discuss the advantages of generating lambda terms in their long normal form and present two novel crossover operations for simply typed lambda terms. Compared to the approach to typed GP over lambda calculus  proposed by Yu\cite{yu01} it uses more general term representation which is not restricting the use of outer local variables in the body of an anonymous function. Compared to the purely combinator approach by Briggs and O’Neill\cite{kes} it has the advantage of reducing the size of the search space in the term generation phase.

@@

The rest of the paper is organized as follows: The next section briefly discusses related work in the field of typed GP, while section~\ref{preliminaries} introduces necessary notions.
The body of the paper is presented in sections \ref{generating} and \ref{crossover}; the first discusses the term generating 
procedure and its relations
with reductions, the second describes several approaches to lambda term
crossover and how they solve the variable problem. Section~\ref{experiments} examines those ideas on the even parity benchmark problem, while the paper is concluded by section~\ref{conclusions}.

@@

Yu presents a GP system utilizing
polymorphic higher-order functions
%\footnote{Higher-order function is a function taking another function as input parameter.} 
and lambda abstractions  \cite{yu01}.
An important point of interest in the work is use of
\texttt{foldr} function as a tool for \textit{implicit recursion}, i.e. recursion without explicit recursive calls. The terminal set for constructing lambda abstraction subtrees  is limited to use only constants and variables of that particular lambda abstraction, i.e., outer variables are not allowed to be used as terminals in this work. This is a significant difference from our approach since we permit all well-typed normalized \lterms. This difference also causes the need for different crossover operators.

Briggs and O’Neill present a technique 
utilizing typed GP with combinators \cite{kes}.
The difference between approach presented in their work
and our approach is that they generate terms in a straightforward way directly from the library of combinators, without any use of lambda abstractions. They use the Hindley–Milner type system. They also present interesting concept of \textit{Generalized genetic operator} based on term generation. The combinator based genetic algorithm has the advantage of avoiding the need of dealing with variables.

Binard and Felty use even stronger type system (\textit{System F}) \cite{binard2008genetic}. But with the increasing power of the type system also comes an increasing difficulty of term generation. For this reason, the evolution in this work takes interesting and nonstandard shape (fitness is associated with \textit{genes} which are evolved together with \textit{species} which together participate in creation of individuals). This differs from our approach, which tries to be a generalization of the standard GP\cite{koza92}.

In contrast with the above mentioned works, our approach uses very simple type system (simply typed lambda calculus) and concentrates on process of generation  
able to generate all possible well-typed normalized lambda terms. In order to do so we use technique based on \textit{inhabitation machines} described by Barendregt\cite{barendregt10}.

@@

Each experiment consisted of 50 independent runs 
of GP algorithm. Each run had maximally 51 generations and population size
was 500 individuals.
In the experiments we analyze the fitness of the best individual,
the average size of an individual,
the ability of the system to produce a correct solution 
and computational cost estimates throughout generations. 
For the last two metrics we use the popular measurement  
methods within the GP field --- the \textit{performance curves}
described in \cite{koza92} --- $P(M,i)$ (cumulative probability of success) 
and $I(M,i,z)$ (the total number of individuals that must be processed 
to yield a correct solution with probability $z =99\%$).
Difference of those two crossover methods is statistically analyzed by the Welch t-test by comparing the fitness values of the best individuals of the run. 
Graphs showing mean values of the fitness of the best individual and
the average size of an individual throughout generations contain error bars representing the standard error of the mean (SEM).

@@

The \textit{unpacking} crossover scored 23/50 (46\%) success rate. 
Minimal $I(M,i,z)$ was in generation 0 with 114,000 individuals to be processed.
The average individual size for generation 50 was 83.1.
The experiment took 388 minutes.

The \textit{hybrid} crossover scored 32/50 (64\%) success rate. 
Minimal $I(M,i,z)$ was in generation 0 with 28,000 individuals to be processed.
The average individual size for generation 50 was 74.3.
The experiment took 33 minutes.

Comparison with other results is summarized in table 2.

@@

We also analyzed ratio of failed crossovers due to exceeding the maximal permissible depth. Only around 0.5\% of offspring created by the unpacking crossover fail due to exceeding the limit and 0.2\% for the hybrid crossover. Thus we conclude that it does not have a significant effect on the performance of the operators.

@@

\begin{figure}[!ht]
  \centering
  \caption{Hybrid crossover example.}
  \includegraphics[scale=0.6]{crossover.png} %0.251
\end{figure}

@@

The first one (\textit{hybrid}) transforms all generated individuals into combinator terms at the end of generation phase. The second crossover approach (\textit{unpacking}) uses the lambda term representation during the whole run, except for the temporary phases during the crossover. We have compared their relative performance on even parity problem; the first one is performing the same or better in all observed metrics. Since the use of the \textit{unpacking} crossover brings a significant computational overhead while its performance is the same or worse than the performance of the \textit{hybrid} crossover we conclude that 
the \textit{hybrid} crossover is better suited for the job.
%\red{This was surprising result for us.}

According to comparison of "individuals to be processed" metric our GP approach is performing better than the GP system presented in \cite{kes} and worse than the one presented in \cite{yu01}. By observation of typical solution which is often some modification of \texttt{foldr xor False inputList} one sees that an important task for GP here is to create \texttt{xor} function. This is the advantage for the more specialized term representation used in \cite{yu01}, which restricts the use of outer variables.

@@

Tomáš Křen has been partially supported by the project GA ČR P202/10/1333 and by the SVV project number 260~104. Roman Neruda has been partially supported by The Ministry of Education of the Czech Republic project COST LD 13002.
